---
title: "01_Cleaning"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true 
    code_folding: hide  
  pdf_document:
    toc: true
    toc_depth: '3' 
output_dir: Output
--- 
```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(janitor)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)


```

# Introduction 

This script will contain the code necessary to clean the data stored in the `/RawData` folder. 

To start it will focus on only cleaning data puled pertaining to the 9000 IMIS classification; later on, subsequent datasets may be cleaned, merged, and analyzed. 

New variables created during the cleaning process, as well as a description of all other included variables, are contained within the excel file `DataDictionary.xlsx`.  

**Goals of Cleaning:** 

1. Ensure that UOM units are consistent 

2. Apply inclusion/exclusion criteria 

3. Add OEL data, create variable that states whether observed sample is above or below.  

# Data Import and Initial Examination

```{r import_data, results= FALSE}
raw_dat <- read_delim("../RawData/Respirable_Silica_Data.txt", delim = "|") 
glimpse(raw_dat)
head(raw_dat) 
tail(raw_dat)
``` 


There are a couple of formatting tasks we need to take care of. We want to convert our data into the proper formats, and there are several variables included in the dataset that we will not need (Office, NAICS, Date_Reported, Lab_Number, CSHO). 

We want to achieve the following: 
- Convert 'result' to be numeric. *Note:* some of these values are recorded as > measurement. To deal with this, we will make a new variable, `result_num` that will report whatever number is associated with the measurement. Another variable, `less_than_measurement`, will indicate whether the variable was "less than" or not.  
- Convert date to an R date format 
- Create a new variable that just contains the year of sample 
- Converting `time` to be in terms of hours, not minutes

Convert result to numeric, check how many are blanks? Also check how many are per sample type and make into nicer format. Check how many went in and used eight hour TWA calculation too. Changing time to be in terms of hours instead of minutes. Adding indicator in case of data being "less than" values, so we can go back in and perhaps do a sensitivity analysis of some kind or just have the knowledge that they were less than that number.

To start, let's simply apply some reformats to our data: 

```{r}
dat <- raw_dat |> 
  rename_with(tolower) |> 
  mutate(
    date_sampled = dmy(date_sampled), 
    year = year(date_sampled),
         type = fct_recode(type,
                          "Personal" = "P",
                          "Area" = "A",
                          "Bulk" = "B", 
                          "Wipe" = "W"), 
    blank = as.logical(blank),
    time = time/60,
    less_than_measurement = grepl("^\\s*<", result),
    result_num = as.numeric(gsub("<\\s*", "", result))
    ) |> 
  select(-office, -naics, -date_reported, -lab_number, -csho, -sic)
```

## Examination  

Let's first see how many years our data spans: 

```{r}
range(dat$date_sampled)
```

We want to check some of these variables to see whether they are relevant to be included (i.e., contain multiple levels). 

Let's check what different values are in the data for `instrument`, `blank`, `weight`, and `uom`. 

```{r}
dat |> 
  group_by(instrument) |> 
  summarize(n()) 
dat |> 
  group_by(blank) |> 
  summarize(n()) 

summary(dat$weight)

dat |> 
  group_by(uom) |> 
  summarize(n()) 
``` 
It looks like we only have NA values for instrument and blank; we can drop. We do have information for weight that may be needed. 

For the unit of measurement, according to the data dictionary, M = mg/m^3, X = micrograms, Y = milligrams. We have 56 NAs which we will need to explore -- it is possible it corresponds to values that contain sample weight and a result, from which we will need to complete some calculations. 

Let's check to see what the NA values correspond to before we reformat the uom variable. 

```{r}
dat |> 
  filter(is.na(uom)) |> 
  select(weight, result_num, qualifier, air_volume)
```

This is interesting, since we have measurements for weight, but we don't know what unit the sample is in for the bulks and silica samples. We are likely unable to use these observations.  

Another important aspect is that we have some units of measurement that are simply micrograms, and milligrams, without being accompanied with the air volume. Let's check and make sure that all of these instances correspond to an air volume value. If they are all complete, we will need to look into converting the air volume (liters) to meters cubed. 

```{r}
dat |> 
  filter(uom == c("M", "MG", "X", "Y")) |> 
  select(uom, air_volume) |>
  group_by(uom) |>
  summarise(
    total_rows = n(),
    non_na_values = sum(!is.na(air_volume)),
    na_count = sum(is.na(air_volume))
  )
```
It appears that we have some data that is missing air volume specifically. We will need to double check which rows are missing air volume after we exclude non-personal samples, and then look into creating a standardized measurement.


Let's apply some reformatting again with the new information.   
 M = mg/m^3, X = micrograms, Y = milligrams.

```{r}
dat <- dat |> 
  select(-instrument, -blank) |> 
  mutate(
    uom = fct_recode(uom,
                          "mg" = "MG", #milligrams
                          "mg" = "Y", #milligrams
                          "mcg" = "X", #micrograms
                          "mcg/m3" = "MCG/M3", #micrograms/m3
                          "mg/m3" = "MG_M3", #milligrams/m3
                          "mg/m3" = "M" #milligrams/m3
                     )
    )
  

dat |> 
  group_by(uom) |> 
  summarize(n())
```



# Inclusion/Exclusion  

## Personal Measurements

To start, we have 2,135 observations. Next, we want to restrict our sample to only those pertaining to personal measurements. 

```{r}
dat |> 
  group_by(type) |> 
  summarize(n())
```

We will retain 1,839 observations that used Personal measurements. We will drop those for Area measurements (n = 20), and missing data (n = 276). 

```{r}
dat <- dat |> 
  filter(type == "Personal")
```

Now that we've excluded some samples, let's go back in and see how the air volume data fits or doesn't: 

```{r}
dat |> 
  filter(uom %in% c("mg", "mcg")) |> 
  select(uom, air_volume) |>
  group_by(uom) |>
  summarise(
    total_rows = n(),
    non_na_values = sum(!is.na(air_volume)),
    na_count = sum(is.na(air_volume))
  )
``` 

Let's also check to see the distribution of samples by year: 

```{r}
dat |> 
  group_by(year) |> 
  summarize(count = n()) |> 
  mutate(percent = (count/sum(count))*100)
```
*Note:* after excluding non-personal samples, we no longer have values that are in milligrams. 

We are left with the resulting microgram observations, which do appear as though they are all lacking data on air volume. 

Let's do one last check, as it is possible that these values correspond to the time-weighted average: 

```{r} 
#TWA groupings for just instances where unit of measurement is mcg
dat |> 
  filter(uom == "mcg") |> 
  select(eighthr) |> 
  group_by(eighthr) |> 
  summarize(n()) 

dat |> 
  filter(uom =="mcg" & eighthr == "N") |> 
  group_by(qualifier) |> 
  summarize(n()) 

dat |> 
  filter(uom =="mcg" & eighthr == "N") |> 
  group_by(air_volume) |>
  summarise(
    total_rows = n(),
    non_na_values = sum(!is.na(air_volume)),
    na_count = sum(is.na(air_volume))
  )

dat |> 
  filter(uom =="mcg" & eighthr == "N") |> 
  select(result_num) |> 
  summary() 

dat |> 
  filter(uom =="mcg" & eighthr == "Y") |> 
  select(result_num) |> 
  summary() 

```
Ok, we have figured out what happened --> specifically the measurements that aren't standardized to volume all contains results less than the quantification limit or are blank.  

Next, we want to scrutinize our observations that are a time-weighted average and are coded as "mcg" for the unit. Let's see what these measurements actually are:  
```{r}
mcg_time_weighted <- dat |> 
  filter(uom =="mcg" & eighthr == "Y")  

mcg_time_weighted |> 
  select(result_num) |> 
  summary() 

#comparing to summary of observations that aren't time-weighted and are coded as "mcg/m3" -- 491 observations
dat |> 
  filter(uom == "mcg/m3" & eighthr == "N") |> 
  select(result_num) |> 
  summary()

#comparing to those that ARE time-weighted and coded as "mcg/m3" -- 737 observations
dat |> 
  filter(uom == "mcg/m3" & eighthr == "Y") |>  
  select(result_num) |> 
  summary()
``` 

We want to go further, tracking if there are overlapping sampling numbers appear across these groups: 
```{r}
mcg_tw    <- dat |> filter(uom == "mcg" & eighthr == "Y")     |> mutate(group = "mcg_TW")
m3_no_tw  <- dat |> filter(uom == "mcg/m3" & eighthr == "N")  |> mutate(group = "m3_notTW")
m3_tw     <- dat |> filter(uom == "mcg/m3" & eighthr == "Y")  |> mutate(group = "m3_TW")

all_three <- bind_rows(mcg_tw, m3_no_tw, m3_tw)

# How many SAMPLING_NUMBERs appear in more than one group?
overlap_check <- all_three |> 
  group_by(sampling_number) |> 
  summarise(
    n_obs = n(),
    n_field_numbers = n_distinct(field_number),
    n_groups = n_distinct(group),
    groups_present = paste(sort(unique(group)), collapse = ", "),
    .groups = "drop"
  ) |> 
  filter(n_groups > 1)

overlap_check

all_three |> 
  arrange(sampling_number, field_number) |> 
  select(sampling_number, field_number, uom, eighthr, result_num, group) |> 
  filter(sampling_number %in% overlap_check$sampling_number) 

all_three |> 
  group_by(field_number) |> 
  summarise(
    n_obs = n(),
    n_groups = n_distinct(group),
    .groups = "drop"
  ) |> 
  filter(n_groups > 1)
```
How many sampling numbers have TWA record:  
```{r}
# does this SAMPLING_NUMBER have an eighthr == "Y" record at all (regardless of uom)?
twa_lookup <- dat |> 
  group_by(sampling_number) |> 
  summarise(
    has_twa = any(eighthr == "Y"),
    n_twa_records = sum(eighthr == "Y"),
    n_component_records = sum(eighthr == "N"),
    .groups = "drop"
  )

twa_lookup |> count(has_twa, n_twa_records)
```

Join back into records to see which are "covered" by a time weighted average:  
```{r}
mcg_components <- dat |> 
  filter(uom == "mcg" & eighthr == "N") |> 
  left_join(twa_lookup, by = "sampling_number")

# how many of these component records belong to a sampling number that DOES have a TWA?
mcg_components |> 
  count(has_twa)
```
Looking at multiple TWA samples
```{r}
dat |> 
  filter(eighthr == "Y") |> 
  group_by(sampling_number) |> 
  filter(n() > 1) |> 
  select(sampling_number, field_number, uom, result_num, time) |>  
  arrange(sampling_number) |> 
  print(n = 30) 


#check for 8-hr calculation: 
check_calc_status <- dat |> 
  filter(eighthr == "Y" & uom == "mcg/m3") |> 
  group_by(sampling_number) |> 
  summarise(
    n_components = n(),
    n_distinct_results = n_distinct(result_num),
    all_results_identical = n_distinct(result_num) == 1,
    total_time = sum(time, na.rm = TRUE),
    .groups = "drop"
  )

# Summary: how many sampling numbers have identical vs differing result_num across components?
check_calc_status |> 
  filter(n_components > 1) |>   # only relevant for sampling numbers with multiple media
  count(all_results_identical) 

# For sampling numbers with >1 component, does a duration-weighted average of the
# individual result_num values recreate any single one of them, or land somewhere in between?
check_calc_status_detail <- dat |> 
  filter(eighthr == "Y" & uom == "mcg/m3") |> 
  group_by(sampling_number) |> 
  filter(n() > 1) |> 
  summarise(
    results = paste(round(result_num, 1), collapse = ", "),
    times = paste(round(time, 2), collapse = ", "),
    computed_twa = sum(result_num * time, na.rm = TRUE) / sum(time, na.rm = TRUE),
    .groups = "drop"
  )

check_calc_status_detail 

check_calc_status2 <- dat |> 
  filter(eighthr == "Y" & uom == "mcg/m3") |> 
  group_by(sampling_number) |> 
  summarise(
    n_components = n(),
    n_distinct_results = n_distinct(result_num),
    all_results_identical = n_distinct(result_num) == 1,
    all_results_zero = all(result_num == 0),
    total_time = sum(time, na.rm = TRUE),
    .groups = "drop"
  )

# Break the "identical" bucket into trivial-zero vs genuine non-zero matches
check_calc_status2 |> 
  filter(n_components > 1) |> 
  count(all_results_identical, all_results_zero)
```

## Removing blank samples 

```{r} 
dat |> 
  group_by(qualifier) |> 
  count() 

dat |> 
  filter(uom == "mcg") |> 
  group_by(qualifier) |> 
  count()

dat <- dat |> 
  filter(!(qualifier %in% c("BLK","ND-BLK", "ND-BLK/t")))
```

## Time-weighted average observations 

Next, we want to get an idea of how our samples may differ depending on how many of them used a time-weighted average. 

```{r}
dat |> 
  group_by(eighthr) |> 
  summarize(n()) 

dat |> 
  filter(eighthr == "STEL CEILING PEAK") |>
  select(everything())  # STEL information 

#examining time distribution by the time-weighted data: 
dat |> 
  group_by(eighthr) |>  
  summarise(
    count = n(),
    mean = mean(result_num, na.rm = TRUE),
    median = median(result_num, na.rm = TRUE),
    min = min(result_num, na.rm = TRUE),
    max = max(result_num, na.rm = TRUE),
    sd = sd(result_num, na.rm = TRUE)
  ) 

#is there a time-trend component to the data that are missing time-weighting? 
dat |> 
  group_by(year, eighthr) |> 
  summarize(count = n(), .groups = 'drop') |>
  ggplot(aes(x = year, y = count, fill = eighthr)) +
  geom_col(position = "dodge") +
  labs(title = "Frequency by Year and Eight Hour Classification",
       x = "Year",
       y = "Count",
       fill = "Eight Hour") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
``` 
Looks like we will not be retaining any of the results corresponding to the STEL ceiling peak as they have way too short times of duration. 


## Over 6-hours of duration 

Let's create an indicator variable for whether the measurements were above 6-hours or not, and take a peek at what data remains: 

```{r} 
dat |> 
  filter(eighthr == "Y") |> 
  count()
dat |> 
  filter(eighthr == "Y") |> 
  count()
dat |> 
  filter(eighthr == "N" & time >=6) |> 
  count() 

dat |> 
  filter(eighthr == "N" & is.na(time) ) |> 
  count()

dat |> 
  filter(eighthr == "N" & time <6) |> 
  count() 

#removing observations that are not time weighted and are below 6 hours in time

dat_filtered <- dat |> 
  filter(!(eighthr == "N" & is.na(time))) |> 
  filter(!(eighthr == "N" & time < 6 & !is.na(time)))

``` 

Before we exclude any data, we want to check if we have repeated measurements per establishments: 

```{r}
 full_sample_check <- dat |>
  group_by(establishment) |>
  summarise(
    n_obs = n(),
    n_years = n_distinct(year),
    .groups = 'drop'
  )

# Summary stats
full_sample_check |>
  summarise(
    total_establishments = n(),
    mean_obs_per_est = mean(n_obs),
    median_obs_per_est = median(n_obs),
    est_with_5plus = sum(n_obs >= 5),
    est_with_10plus = sum(n_obs >= 10),
    pct_5plus = round(100 * sum(n_obs >= 5) / n(), 1),
    pct_10plus = round(100 * sum(n_obs >= 10) / n(), 1)
  )  

# Distribution
full_sample_check |>
  ggplot(aes(x = n_obs)) +
  geom_histogram(binwidth = 2, fill = "steelblue") +
  geom_vline(xintercept = c(5, 10), linetype = "dashed", color = "red") +
  labs(title = "Distribution of Observations per Establishment (Full Sample)",
       x = "Number of Observations",
       y = "Number of Establishments") +
  theme_minimal() 

#Restricted sample: 
restricted_sample_check <- dat |>
  filter(six_hour == 1) |>  # or your > 6 hour filter
  group_by(establishment) |>
  summarise(
    n_obs = n(),
    n_years = n_distinct(year),
    .groups = 'drop'
  )

restricted_sample_check |>
  summarise(
    total_establishments = n(),
    mean_obs_per_est = mean(n_obs),
    median_obs_per_est = median(n_obs),
    est_with_5plus = sum(n_obs >= 5),
    est_with_10plus = sum(n_obs >= 10),
    pct_5plus = round(100 * sum(n_obs >= 5) / n(), 1),
    pct_10plus = round(100 * sum(n_obs >= 10) / n(), 1)
  ) 

#summary table 
tribble(
  ~criterion, ~guideline, ~full_sample, ~restricted_sample,
  "Number of establishments", "≥20-30", nrow(full_sample_check), nrow(restricted_sample_check),
  "% with 5+ observations", "~80%+", 
  round(100 * sum(full_sample_check$n_obs >= 5) / nrow(full_sample_check), 1),
  round(100 * sum(restricted_sample_check$n_obs >= 5) / nrow(restricted_sample_check), 1),
  "% with 10+ observations", "~50%+",
  round(100 * sum(full_sample_check$n_obs >= 10) / nrow(full_sample_check), 1),
  round(100 * sum(restricted_sample_check$n_obs >= 10) / nrow(restricted_sample_check), 1)
)
``` 

We do not meet the assumptions of LMEM according to [this article](https://ladal.edu.au/tutorials/mixedmodel/mixedmodel.html), which details that we need 5-10 observations per random-effect level for reliable variance estimates, and at least 20-30 random-effect levels. 

It is recommended that no matter what, we apply cluster-robust standard errors to account for within-group variation. To deal with this, we are going to plan on fitting a generalized linear mixed model that adds a cluster-specific random effect to the linear predictor. Our cluster variable will be id. 


