---
title: "01_Cleaning"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true 
    code_folding: hide  
  pdf_document:
    toc: true
    toc_depth: '3' 
output_dir: Output
--- 
```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(janitor)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)


```

# Introduction 

This script will contain the code necessary to clean the data stored in the `/RawData` folder. 

To start it will focus on only cleaning data puled pertaining to the 9000 IMIS classification; then it will generate the samples pertaining to 9130 (respirable dust concentration) and 9010 (crystalline quartz fraction [%]) to calculate silica concentrations for years prior to the classification of 9000. 

New variables created during the cleaning process, as well as a description of all other included variables, are contained within the excel file `DataDictionary.xlsx`.  

**Goals of Cleaning:** 

1. Ensure that UOM units are consistent 

2. Apply inclusion/exclusion criteria 

3. Add OEL data, create variable that states whether observed sample is above or below.   

#IMIS 9000 data

## Data Import and Initial Examination

```{r import_data, results= FALSE}
raw_dat <- read_delim("../RawData/Respirable_Silica_Data.txt", delim = "|") 
glimpse(raw_dat)
head(raw_dat) 
tail(raw_dat)
``` 


There are a couple of formatting tasks we need to take care of. We want to convert our data into the proper formats, and there are several variables included in the dataset that we will not need (Office, NAICS, Date_Reported, Lab_Number, CSHO). 

We want to achieve the following: 
- Convert 'result' to be numeric. *Note:* some of these values are recorded as > measurement. To deal with this, we will make a new variable, `result_num` that will report whatever number is associated with the measurement. Another variable, `less_than_measurement`, will indicate whether the variable was "less than" or not.  
- Convert date to an R date format 
- Create a new variable that just contains the year of sample 
- Converting `time` to be in terms of hours, not minutes

Convert result to numeric, check how many are blanks? Also check how many are per sample type and make into nicer format. Check how many went in and used eight hour TWA calculation too. Changing time to be in terms of hours instead of minutes. Adding indicator in case of data being "less than" values, so we can go back in and perhaps do a sensitivity analysis of some kind or just have the knowledge that they were less than that number.

To start, let's simply apply some reformats to our data. We will make this a function to save time later: 

```{r} 
clean_osha <- function(df){
  df |> 
    rename_with(tolower) |> 
  mutate(
    date_sampled = dmy(date_sampled), 
    year = year(date_sampled), 
         type = fct_recode(type,
                          "Personal" = "P",
                          "Area" = "A",
                          "Bulk" = "B", 
                          "Wipe" = "W"),  
    uom = fct_recode(uom,
                          "mg" = "MG", #milligrams
                          "mg" = "Y", #milligrams
                          "mcg" = "X", #micrograms
                          "mcg/m3" = "MCG/M3", #micrograms/m3
                          "mg/m3" = "MG_M3", #milligrams/m3
                          "mg/m3" = "M"), #milligrams/m3
    blank = as.logical(blank),
    time = time/60,
    less_than_measurement = grepl("^\\s*<", result),
    result_num = as.numeric(gsub("<\\s*", "", result))
    ) |> 
  select(-office, -naics, -date_reported, -csho, -sic, -instrument, -blank) 
}
dat <- clean_osha(raw_dat) 
  
```

### Examination  

Let's first see how many years our data spans: 

```{r}
range(dat$date_sampled)
```

We want to check some of these variables to see whether they are relevant to be included (i.e., contain multiple levels). 


It looks like we only have NA values for instrument and blank; we can drop. We do have information for weight that may be needed. 

For the unit of measurement, according to the data dictionary, M = mg/m^3, X = micrograms, Y = milligrams. We have 56 NAs which we will need to explore -- it is possible it corresponds to values that contain sample weight and a result, from which we will need to complete some calculations. 

Let's check to see what the NA values correspond to before we reformat the uom variable. 

```{r}
dat |> 
  filter(is.na(uom)) |> 
  select(weight, result_num, qualifier, air_volume)
```

This is interesting, since we have measurements for weight, but we don't know what unit the sample is in for the bulks and silica samples. We are likely unable to use these observations.  

Another important aspect is that we have some units of measurement that are simply micrograms, and milligrams, without being accompanied with the air volume. Let's check and make sure that all of these instances correspond to an air volume value. If they are all complete, we will need to look into converting the air volume (liters) to meters cubed. 

```{r}
dat |> 
  filter(uom == c("M", "MG", "X", "Y")) |> 
  select(uom, air_volume) |>
  group_by(uom) |>
  summarise(
    total_rows = n(),
    non_na_values = sum(!is.na(air_volume)),
    na_count = sum(is.na(air_volume))
  )
```
It appears that we have some data that is missing air volume specifically. We will need to double check which rows are missing air volume after we exclude non-personal samples, and then look into creating a standardized measurement.


Let's apply some reformatting again with the new information.   
 M = mg/m^3, X = micrograms, Y = milligrams.

```{r}
dat |> 
  group_by(uom) |> 
  summarize(n())
```

### Personal Measurements

To start, we have 2,135 observations. Next, we want to see how many measurements pertain to personal measurements. 

```{r}
dat |> 
  group_by(type) |> 
  summarize(n())
```



Let's also check to see the distribution of samples by year: 

```{r}
dat |> 
  group_by(year) |> 
  summarize(count = n()) |> 
  mutate(percent = (count/sum(count))*100)
```
*Note:* after excluding non-personal samples, we no longer have values that are in milligrams. 

We are left with the resulting microgram observations, which do appear as though they are all lacking data on air volume. 

Let's do one last check, as it is possible that these values correspond to the time-weighted average: 

```{r} 
#TWA groupings for just instances where unit of measurement is mcg
dat |> 
  filter(uom == "mcg") |> 
  select(eighthr) |> 
  group_by(eighthr) |> 
  summarize(n()) 

dat |> 
  filter(uom =="mcg" & eighthr == "N") |> 
  group_by(qualifier) |> 
  summarize(n()) 

dat |> 
  filter(uom =="mcg" & eighthr == "N") |> 
  group_by(air_volume) |>
  summarise(
    total_rows = n(),
    non_na_values = sum(!is.na(air_volume)),
    na_count = sum(is.na(air_volume))
  )

dat |> 
  filter(uom =="mcg" & eighthr == "N") |> 
  select(result_num) |> 
  summary() 

dat |> 
  filter(uom =="mcg" & eighthr == "Y") |> 
  select(result_num) |> 
  summary() 

```
Ok, we have figured out what happened --> specifically the measurements that aren't standardized to volume all contains results less than the quantification limit or are blank.  

Next, we want to scrutinize our observations that are a time-weighted average and are coded as "mcg" for the unit. Let's see what these measurements actually are:  
```{r}
mcg_time_weighted <- dat |> 
  filter(uom =="mcg" & eighthr == "Y")  

mcg_time_weighted |> 
  select(result_num) |> 
  summary() 

#comparing to summary of observations that aren't time-weighted and are coded as "mcg/m3" -- 491 observations
dat |> 
  filter(uom == "mcg/m3" & eighthr == "N") |> 
  select(result_num) |> 
  summary()

#comparing to those that ARE time-weighted and coded as "mcg/m3" -- 737 observations
dat |> 
  filter(uom == "mcg/m3" & eighthr == "Y") |>  
  select(result_num) |> 
  summary()
``` 

We want to go further, tracking if there are overlapping sampling numbers appear across these groups: 
```{r}
mcg_tw    <- dat |> filter(uom == "mcg" & eighthr == "Y")     |> mutate(group = "mcg_TW")
m3_no_tw  <- dat |> filter(uom == "mcg/m3" & eighthr == "N")  |> mutate(group = "m3_notTW")
m3_tw     <- dat |> filter(uom == "mcg/m3" & eighthr == "Y")  |> mutate(group = "m3_TW")

all_three <- bind_rows(mcg_tw, m3_no_tw, m3_tw)

# How many SAMPLING_NUMBERs appear in more than one group?
overlap_check <- all_three |> 
  group_by(sampling_number) |> 
  summarise(
    n_obs = n(),
    n_field_numbers = n_distinct(field_number),
    n_groups = n_distinct(group),
    groups_present = paste(sort(unique(group)), collapse = ", "),
    .groups = "drop"
  ) |> 
  filter(n_groups > 1)

overlap_check

all_three |> 
  arrange(sampling_number, field_number) |> 
  select(sampling_number, field_number, uom, eighthr, result_num, group) |> 
  filter(sampling_number %in% overlap_check$sampling_number) 

all_three |> 
  group_by(field_number) |> 
  summarise(
    n_obs = n(),
    n_groups = n_distinct(group),
    .groups = "drop"
  ) |> 
  filter(n_groups > 1)
```
How many sampling numbers have TWA record:  
```{r}
# does this SAMPLING_NUMBER have an eighthr == "Y" record at all (regardless of uom)?
twa_lookup <- dat |> 
  group_by(sampling_number) |> 
  summarise(
    has_twa = any(eighthr == "Y"),
    n_twa_records = sum(eighthr == "Y"),
    n_component_records = sum(eighthr == "N"),
    .groups = "drop"
  )

twa_lookup |> count(has_twa, n_twa_records)
```

Join back into records to see which are "covered" by a time weighted average:  
```{r}
mcg_components <- dat |> 
  filter(uom == "mcg" & eighthr == "N") |> 
  left_join(twa_lookup, by = "sampling_number")

# how many of these component records belong to a sampling number that DOES have a TWA?
mcg_components |> 
  count(has_twa)
```
Looking at multiple TWA samples
```{r}
dat |> 
  filter(eighthr == "Y") |> 
  group_by(sampling_number) |> 
  filter(n() > 1) |> 
  select(sampling_number, field_number, uom, result_num, time) |>  
  arrange(sampling_number) |> 
  print(n = 30) 


#check for 8-hr calculation: 
check_calc_status <- dat |> 
  filter(eighthr == "Y" & uom == "mcg/m3") |> 
  group_by(sampling_number) |> 
  summarise(
    n_components = n(),
    n_distinct_results = n_distinct(result_num),
    all_results_identical = n_distinct(result_num) == 1,
    total_time = sum(time, na.rm = TRUE),
    .groups = "drop"
  )

# Summary: how many sampling numbers have identical vs differing result_num across components?
check_calc_status |> 
  filter(n_components > 1) |>   # only relevant for sampling numbers with multiple media
  count(all_results_identical) 

# For sampling numbers with >1 component, does a duration-weighted average of the
# individual result_num values recreate any single one of them, or land somewhere in between?
check_calc_status_detail <- dat |> 
  filter(eighthr == "Y" & uom == "mcg/m3") |> 
  group_by(sampling_number) |> 
  filter(n() > 1) |> 
  summarise(
    results = paste(round(result_num, 1), collapse = ", "),
    times = paste(round(time, 2), collapse = ", "),
    computed_twa = sum(result_num * time, na.rm = TRUE) / sum(time, na.rm = TRUE),
    .groups = "drop"
  )

check_calc_status_detail 

check_calc_status2 <- dat |> 
  filter(eighthr == "Y" & uom == "mcg/m3") |> 
  group_by(sampling_number) |> 
  summarise(
    n_components = n(),
    n_distinct_results = n_distinct(result_num),
    all_results_identical = n_distinct(result_num) == 1,
    all_results_zero = all(result_num == 0),
    total_time = sum(time, na.rm = TRUE),
    .groups = "drop"
  )

# Break the "identical" bucket into trivial-zero vs genuine non-zero matches
check_calc_status2 |> 
  filter(n_components > 1) |> 
  count(all_results_identical, all_results_zero)
```



### Time-weighted average observations 

Next, we want to get an idea of how our samples may differ depending on how many of them used a time-weighted average. 

```{r}
dat |> 
  group_by(eighthr) |> 
  summarize(n()) 

dat |> 
  filter(eighthr == "STEL CEILING PEAK") |>
  select(everything())  # STEL information 

#examining time distribution by the time-weighted data: 
dat |> 
  group_by(eighthr) |>  
  summarise(
    count = n(),
    mean = mean(result_num, na.rm = TRUE),
    median = median(result_num, na.rm = TRUE),
    min = min(result_num, na.rm = TRUE),
    max = max(result_num, na.rm = TRUE),
    sd = sd(result_num, na.rm = TRUE)
  ) 

#is there a time-trend component to the data that are missing time-weighting? 
dat |> 
  group_by(year, eighthr) |> 
  summarize(count = n(), .groups = 'drop') |>
  ggplot(aes(x = year, y = count, fill = eighthr)) +
  geom_col(position = "dodge") +
  labs(title = "Frequency by Year and Eight Hour Classification",
       x = "Year",
       y = "Count",
       fill = "Eight Hour") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
``` 
Looks like we will not be retaining any of the results corresponding to the STEL ceiling peak as they have way too short times of duration. 


### Over 6-hours of duration 

Let's create an indicator variable for whether the measurements were above 6-hours or not, and take a peek at what data remains: 

```{r} 
dat |> 
  filter(eighthr == "Y") |> 
  count()
dat |> 
  filter(eighthr == "Y") |> 
  count()
dat |> 
  filter(eighthr == "N" & time >=6) |> 
  count() 

dat |> 
  filter(eighthr == "N" & is.na(time) ) |> 
  count()

dat |> 
  filter(eighthr == "N" & time <6) |> 
  count() 

``` 

Before we exclude any data, we want to check if we have repeated measurements per establishments: 

```{r}
 full_sample_check <- dat |>
  group_by(establishment) |>
  summarise(
    n_obs = n(),
    n_years = n_distinct(year),
    .groups = 'drop'
  )

# Summary stats
full_sample_check |>
  summarise(
    total_establishments = n(),
    mean_obs_per_est = mean(n_obs),
    median_obs_per_est = median(n_obs),
    est_with_5plus = sum(n_obs >= 5),
    est_with_10plus = sum(n_obs >= 10),
    pct_5plus = round(100 * sum(n_obs >= 5) / n(), 1),
    pct_10plus = round(100 * sum(n_obs >= 10) / n(), 1)
  )  

# Distribution
full_sample_check |>
  ggplot(aes(x = n_obs)) +
  geom_histogram(binwidth = 2, fill = "steelblue") +
  geom_vline(xintercept = c(5, 10), linetype = "dashed", color = "red") +
  labs(title = "Distribution of Observations per Establishment (Full Sample)",
       x = "Number of Observations",
       y = "Number of Establishments") +
  theme_minimal() 

#Restricted sample: 
restricted_sample_check <- dat |>
  filter(time >= 6) |>  # or your > 6 hour filter
  group_by(establishment) |>
  summarise(
    n_obs = n(),
    n_years = n_distinct(year),
    .groups = 'drop'
  )

restricted_sample_check |>
  summarise(
    total_establishments = n(),
    mean_obs_per_est = mean(n_obs),
    median_obs_per_est = median(n_obs),
    est_with_5plus = sum(n_obs >= 5),
    est_with_10plus = sum(n_obs >= 10),
    pct_5plus = round(100 * sum(n_obs >= 5) / n(), 1),
    pct_10plus = round(100 * sum(n_obs >= 10) / n(), 1)
  ) 

#summary table 
tribble(
  ~criterion, ~guideline, ~full_sample, ~restricted_sample,
  "Number of establishments", "≥20-30", nrow(full_sample_check), nrow(restricted_sample_check),
  "% with 5+ observations", "~80%+", 
  round(100 * sum(full_sample_check$n_obs >= 5) / nrow(full_sample_check), 1),
  round(100 * sum(restricted_sample_check$n_obs >= 5) / nrow(restricted_sample_check), 1),
  "% with 10+ observations", "~50%+",
  round(100 * sum(full_sample_check$n_obs >= 10) / nrow(full_sample_check), 1),
  round(100 * sum(restricted_sample_check$n_obs >= 10) / nrow(restricted_sample_check), 1)
)
``` 

We do not meet the assumptions of LMEM according to [this article](https://ladal.edu.au/tutorials/mixedmodel/mixedmodel.html), which details that we need 5-10 observations per random-effect level for reliable variance estimates, and at least 20-30 random-effect levels. 

It is recommended that no matter what, we apply cluster-robust standard errors to account for within-group variation. To deal with this, we are going to plan on fitting a generalized linear mixed model that adds a cluster-specific random effect to the linear predictor. Our cluster variable will be id. 
 
## Unit of Measurement Consistency 

We need to make sure we are only using reported results under one standard unit of measurement. 

Since the OSHA limits correspond to micrograms/m^3, we will convert our other measurement to be consistent with this.  

Just to review, let's double-check what the possible values are for our silica measurements: 

```{r}
dat |> 
  group_by(uom) |> 
  summarize(n())
```
We have 56 that are NA, let's just double check that these aren't all time-weighted before we consider deleting them: 

```{r}
dat |> 
  filter(is.na(uom)) |> 
  group_by(eighthr) |> 
  summarize(n())
```
There are 39 observations that are not time-weighted, so we need to be sure to exclude them from the sample. We are also going to plan on doing a sensitivity analysis on the observations that have no uom but are time-weighted; we cannot be sure what units these are in, so we can add them back in later and see whether things change significantly.  

### Comparing mg to mcg 

Before we convert all our units, let's do a check to see how the different concentrations differ: 

```{r}
dat |> 
  group_by(uom) |> 
  summarize(n= n(),
            mean = mean(result_num),
            sd = sd(result_num))
```
Looks like we need to see how many observations are missing data for the mg/m3 observations: 
```{r}
dat |> 
  filter(uom == "mg/m3") |> 
  summarize(n = sum(is.na(result_num)))

dat |> 
  filter(uom == "mg/m3" ) |> 
  group_by(qualifier) |> 
  summarize(n())
```
## Establishment standardization

Now that we have our datasets for respirable silica and our calculated versions, we need to prepare for the merge by deleting duplicate entries. We have some duplicates that are the same field and lab number, due to different spellings of the establishment.  

We are going to make a function that resolves duplicate lab_number + field_number entries caused by inconsistent spelling.


```{r}
standardize_establishment <- function(x) {
  x |>
    str_squish() |>                          # trim + collapse internal whitespace
    str_to_upper() |>
    str_replace_all("&", "AND") |>           # unify ampersand vs "AND"
    str_replace_all("[[:punct:]]", "") |>    # strip periods, commas, hyphens, etc.
    str_squish()                             # re-squish in case punctuation removal left double spaces
} 


dat <- dat |> mutate(establishment = standardize_establishment(establishment)) 

```



## Inclusion/Exclusion    

To begin, let's see how many observations we have: 

```{r}
nrow(dat)
```

We are starting with 2,135.

### Correct Measurement   


We need our data to record a legitimate recorded measurement. First, we will check to see what the samples are recorded as when they are listed as being LOQ:

```{r} 

dat |>  
  summarise(sum(is.na(result_num)))

dat |> 
  filter(is.na(result_num)) |> 
  group_by(qualifier) |> 
  summarise(n()) 

```
 There are 85 observations that are missing data for the recorded measurement. As such we will have to delete these observations. 
 
```{r} 
dat2 <- dat |> 
  filter(!is.na(result_num))
```
 After removal, we now have 2,050 observations. 
 
### Removing samples lacking concentration data

We are only going to retain samples that contain clear units standardized to air volume. We cannot convert our observations to a concentration that are standalones, since they lack air volume data. 

To our best knowledge, the records cannot be personal air samples, and were likely bulk samples that were miscoded as personal samples.

```{r}
dat3 <- dat2 |> 
    mutate(
    result_num = case_when(
      uom == "mg" ~ result_num * 1000,      # mg → mcg (multiply by 1000)
      uom == "mg/m3" ~ result_num * 1000,   # mg/m3 → mcg/m3 (multiply by 1000)
      TRUE ~ result_num
    ),
    uom = case_when(
      uom == "mg" ~ "mcg",
      uom == "mg/m3" ~ "mcg/m3",
      TRUE ~ uom
    )
  ) 

dat3 |> 
  group_by(uom) |> 
  summarise(n())
```



At this time, there are 572 measurements that are not in concentration form. We will exclude: 

```{r}
dat3 <- dat3 |> 
  filter(uom == "mcg/m3")
```

We now have 1,396 observations. We will rename the results into a consistent variable name for later, called "silica_conc": 

```{r}
dat3 <- dat3 |> 
  mutate(silica_conc = result_num)
```


### No blank samples 
Next we will exclude all samples that were recorded as blank. 

```{r} 
dat3 |> 
  group_by(qualifier) |> 
  count() 


dat4 <- dat3 |> 
  filter(!(qualifier %in% c("BLK","ND-BLK", "ND-BLK/t")))
```
We gave removed blank samples (n =1) and samples that were recorded as both ND (LOQ) and Blank (n = 368), and those recorded as ND, Blank/t (n =3).  

We now have 1470 observations.

### No duplicate samples   

We need to ensure that we don't have duplicate samples in the data.  

```{r}
nrow(dat4) #how many rows in data?  

#look at duplicates 
dat4_duplicates <- dat4 |>
  group_by(lab_number, field_number, silica_conc, date_sampled) |>
  filter(n() > 1) |>
  ungroup() 

dat4_duplicates |>  
  select(lab_number, field_number, date_sampled, establishment, silica_conc) |> 
  head()

#all variables 
dat4_duplicates |>   
  head()

dat4_dedup <- dat4|> 
  distinct(lab_number, field_number, silica_conc, date_sampled, 
           .keep_all = TRUE) 

nrow(dat4_dedup) #how many rows are left after deleting duplicates? 


```

### Only Personal Measurements  

We only want to retain personal measurements for this analysis: 

```{r}
dat4_dedup |> 
  group_by(type) |> 
  count() 

dat5 <- dat4_dedup |> 
  filter(type == "Personal")
```

We are left with 1,299 observations.


### >= 6 hrs

We need to ensure that we have measurements that were taken over 6 hours in length:

```{r} 
dat5 |> 
  mutate(six_hours = time >= 6) |>
  filter(six_hours == "TRUE") |> 
  group_by( year) |> 
  summarise(n()) #checking that we have enough observations

dat5 |> 
  filter(is.na(time)) |> 
  count() 
dat5 |> 
  filter(time < 6) |> 
  count() 

dat5 |> 
  filter(time < 6 & is.na(time)) |> 
  count() 

dat6 <- dat5 |> 
  filter(!(time < 6 & !is.na(time)))
``` 
6 are missing time data. n = 797 are less than 6 hours in duration.   

After excluding, we are left with 599 observations.


Let's check and see how many observations we have: 

```{r}
glimpse(dat6)
``` 

Let's clean up our data too, only retaining the variables we need: 

```{r}
dat7 <- dat6 |> 
  select(sampling_number, establishment, date_sampled, lab_number, field_number, time, imis, year, silica_conc)
```


# IMIS 9130 and 9010 Data Import 

Next, we need to import the data pertaining to the other two codes. 
We need to be certain that the samples match with each other (matching on "Lab No."), before multiplying the respirable dust concentration by the quartz percentage for corresponding samples.  

```{r}
dust_dat <- read_delim("../RawData/Respirable_Dust_Concentration.txt", delim = "|") 
glimpse(dust_dat)
head(dust_dat) 
tail(dust_dat) 

quartz_dat <- read_delim("../RawData/Crystalline_Quartz_Fraction.txt", delim = "|") 
glimpse(quartz_dat)
head(quartz_dat) 
tail(quartz_dat)
```

## Reformatting using Clean_OSHA Function

We can reuse our clean_osha function here to have consistent column naming and cleaning rules:

```{r}
dust2 <- clean_osha(dust_dat) 
  
dust2 |> 
  group_by(uom) |> 
  summarize(n()) 

range(dust2$year)

quartz2 <-clean_osha(quartz_dat)
 

```

## Filter Quartz for percentage measurements 

We need to double check what the units of measurement and recorded numbers look like in this data. 

Fortunately, the variable for result is numeric, so we don't need to worry about the "<" measurements. 

```{r} 
nrow(quartz2)
quartz2 |> 
  group_by(uom) |> 
  count() 
```
"AAAAA" likely refers to the samples that had blanks or other problems.

For our quartz data specifically, we want to retain only quartz fraction measurements expressed as percentages (%). Exclude measurements in other units (mcg, mg, invalid codes) as these are not quartz fractions.

Inclusion Criteria: Unit of Measurement (uom) == "%"

Exclusion Reasons:

AAAAA: Blanks or quality control issues
N: Non-numeric/invalid codes
mcg, mg: Wrong units for quartz fraction
NA: Missing units 

```{r}
nrow(quartz2) 

quartz3 <- quartz2 |> 
  filter(uom == "%") 

nrow(quartz3)
```
We started with 4,672 observations, and excluded observations not pertaining to a % (AAAAA: n =11, N: n = 2, mcg: n = 326, mg: n = 20, NA: n = 136). We now have 4,177 observations. 

We can also make sure to convert our result over to a percentage right now: 

```{r}
quartz3 <- quartz3 |> 
  mutate(quartz_pct = result_num/100)
```


## Filter Dust for Respirable Concentration in mg/m3

We need to retain only respirable dust measurements expressed in mg/m³. Convert to mcg/m³ for consistency with other exposure metrics.

**Exclusion Reasons:**

AAAAA: Quality control samples
mg: Absolute mass (not concentration)
mcg: Wrong unit for this measurement
NA: Missing unit information 


We know that the result for dust is similar to the respirable silica dataset that included "<" measurements, so we will also need to create a numeric version of the results, making note of this in our results.

```{r}
dust2 |> 
  group_by(uom) |> 
  count()
```
We are going to retain the data coded as mg/m3 as the measured concentration. Accordingly, we will subset our data: 

```{r}
nrow(dust2) 

dust3 <- dust2 |> 
  filter(uom == "mg/m3") 

nrow(dust3)
```

We started with 3,893 observations, and excluded units of measurement of AAAAA (n =2), mg (n = 251), mcg (n =1), and NAs (n = 5). We now have 3,633 observations.  

###Convert to mcg/m3

We are going to do our manual conversion to mcg from mg right now to keep things consistent: 

```{r}
dust3 <- dust3 |> 
  mutate(dust_conc = result_num*1000,
         dust_uom = "mcg/m3") 
```

##Establishment standard 

```{r}
dust3 <- dust3 |> mutate(establishment = standardize_establishment(establishment))
quartz3 <- quartz3 |> mutate(establishment = standardize_establishment(establishment))
```





## Remove Blanks 

We need to ensure that we don't have blank measurements in either the dust or quartz datasets: 

```{r}
dust3 |> 
  group_by(qualifier) |> 
  count()

quartz3|> 
  group_by(qualifier) |> 
  count() 

dust4 <- dust3 |> 
  filter(!(qualifier %in% c("BLK","ND-BLK", "ND-BLK/t", "ND-BLK\t\t\t", "ND-BLK\t\t\t\t"))) 

quartz4 <- quartz3 |> 
  filter(!(qualifier %in% c("BLK","ND-BLK", "ND-BLK/t", "ND-BLK\t\t\t", "ND-BLK\t\t\t\t")))

```


## No duplicate samples 


We only want to retain values that have the same Lab No., and values that correspond to the same date sampled. We are going to join these two datasets based on lab_number and date_sampled to be sure that the joins correspond to the same day/sample. First, we are going to rename our results from these two datasets into "dust_conc" and "quartz_pct". 

```{r}
dust4 <- dust4 |> 
  select(lab_number, date_sampled, field_number, sampling_number, 
         establishment, type, eighthr, time, dust_conc, dust_uom)

quartz4 <- quartz4 |> 
  select(lab_number, date_sampled, field_number, sampling_number, establishment,
          quartz_pct)
```

Checks before joining: 

Does either table have duplicate lab_number + date_sampled + establishment + field_number + result combos?  

Dust first:

###Dust

```{r}

#look at duplicates dust
dust4_duplicates <- dust4 |>
  group_by(lab_number, field_number, dust_conc, date_sampled) |>
  filter(n() > 1) |>
  ungroup() 

dust4_duplicates |>  
  arrange(lab_number, field_number, date_sampled, establishment, dust_conc) |> 
  select(lab_number, field_number, date_sampled, establishment, dust_conc) |> 
  head()

#all variables 
dust4_duplicates |> 
  arrange(lab_number, field_number, date_sampled, establishment, dust_conc) |> 
  head()

dust4_dedup <- dust4|> 
  distinct(lab_number, field_number, dust_conc, date_sampled, 
           .keep_all = TRUE) 

nrow(dust4_dedup) #how many rows are left after deleting duplicates?  
``` 

###Quartz 

Next, quartz: 


```{r}
#look at duplicates quartz
quartz4_duplicates <- quartz4 |>
  group_by(lab_number, field_number, quartz_pct, date_sampled) |>
  filter(n() > 1) |>
  ungroup() 

quartz4_duplicates |>  
  arrange(lab_number, field_number, date_sampled, establishment, quartz_pct) |> 
  select(lab_number, field_number, date_sampled, establishment, quartz_pct) |> 
  head()

#all variables 
quartz4_duplicates |>   
  arrange(lab_number, field_number, date_sampled, establishment, quartz_pct) |> 
  head()

quartz4_dedup <- quartz4|> 
  distinct(lab_number, field_number, quartz_pct, date_sampled, 
           .keep_all = TRUE) 

nrow(quartz4_dedup) #how many rows are left after deleting duplicates? 
```



**NOTE**: when we performed this, we lost some information about different spellings of the establishments. This will likely come up again as being problematic when we merge with the other respirable silica dataset. Something to be aware of in the future. 


## Removing unnecessary columns

We are only going to retain the join keys and quartz pct before merging with dust: 

```{r} 

# Remove unnecessary columns
quartz_merge <- quartz4_dedup |> 
  select(lab_number, field_number, quartz_pct)

#removing the hyphen part after number
dust_merge <- dust4_dedup |> 
  mutate(sampling_number = word(sampling_number, 1, sep = "-"))  

glimpse(dust_merge) 
glimpse(quartz_merge) 


``` 


```{r}

merged_dust_quartz <- quartz_merge|> 
  inner_join(dust_merge, by = c("lab_number", "field_number"),
             relationship = "one-to-one")  

#checking how many didn't match: 
quartz_merge|> 
  anti_join(dust_merge, by = c("lab_number", "field_number")) |> 
  nrow()

nrow(merged_dust_quartz) 
glimpse(merged_dust_quartz)
``` 
Ok, looks like we have 1211 observations.  

Let's perform our calculation to get the official silica concentration, and add a variable for the year: 

```{r}
merged_dust_quartz2 <- merged_dust_quartz |> 
  mutate(silica_conc = dust_conc*quartz_pct,
         year = year(date_sampled)) 

merged_dust_quartz2 |> 
  summarize(mean = mean(silica_conc),sd = sd(silica_conc)) 

#by year: 
merged_dust_quartz2 |> 
  group_by(year) |> 
  summarize(mean = mean(silica_conc),sd = sd(silica_conc)) 
```



##Personal Measures 

We need to ensure that we are only retaining measurements that are personal:

```{r}
merged_dust_quartz2 |> 
  group_by(type) |> 
  count() 

merged_dust_quartz3 <- merged_dust_quartz2|> 
  filter(type == "Personal")
```


## >= 6 hrs 

```{r} 
merged_dust_quartz3 |> 
  mutate(six_hours = time >= 6) |>
  filter(six_hours == "TRUE") |> 
  group_by( year) |> 
  summarise(n()) #checking that we have enough observations

merged_dust_quartz3|> 
  filter(is.na(time)) |> 
  count() 
merged_dust_quartz3|> 
  filter(time < 6) |> 
  count() 

merged_dust_quartz3 |> 
  filter(time < 6 & is.na(time)) |> 
  count() 

merged_dust_quartz4 <- merged_dust_quartz3|> 
  filter(!(time < 6 & !is.na(time)))
``` 

# Merge Respirable and Calculated data 

Lastly, we need to merge our quartz and dust datasets with the respirable dataset. Before we do this, we need to make sure we include only the relevant variables, as well as make a consistent respirable concentration variable that only has results in mcg/m3.  


Variables to add from merged_dust_quarts: imis (val = 9010_9130) 
To drop: quartz_pct, dust_conc, dust_uom. 

```{r}
merged_dust_quartz5 <- merged_dust_quartz4 |> 
  select(-quartz_pct, -dust_conc, -dust_uom) |> 
  mutate(imis = "9010_9130") 

dat7 <- dat7 |> 
  mutate(imis = as.character(imis))
```

```{r}
combined_dat <- bind_rows(dat7, merged_dust_quartz5)
nrow(combined_dat) 
glimpse(combined_dat)
``` 

#Data check 

We want to see how many observations we have per year, as well as check for potential outliers: 

```{r}
summary(combined_dat$silica_conc)
Q1 <- quantile(combined_dat$silica_conc, 0.25, na.rm = TRUE)
Q3 <- quantile(combined_dat$silica_conc, 0.75, na.rm = TRUE)
IQR_value <- IQR(combined_dat$silica_conc, na.rm = TRUE)

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

lower_bound
upper_bound 
outliers <- combined_dat |> 
  filter(silica_conc < lower_bound | silica_conc > upper_bound)

outliers 

combined_dat |> 
  slice_max(silica_conc, n = 5)
```
For our outliers, we want to be certain that none of these started as a different unit of measurement. We will compare with our original database: 

```{r}
outlier_field_no <- combined_dat |> 
  slice_max(silica_conc, n = 5) |> 
  pull(field_number)  

outlier_field_no 

dat |> 
  filter(field_number %in% outlier_field_no) |> 
  select(uom, result_num, eighthr) 

combined_dat |> 
  group_by(eighthr) |> 
  count()
```
Ok, all of these started off as mcg/m3 interestingly. We cannot chalk it up to being incorrect measurement classification.

Let's save our analytic dataset: 

```{r}
write_csv(combined_dat, "../CleanData/study_dat.csv")
```