---
title: "02_Exploration"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true 
    code_folding: hide 
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(janitor) 
library(viridis) 
library(sandwich)
library(lmtest) 
library(broom)


# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

study_dat <- read.csv("../CleanData/study_dat.csv") 

study_dat <- study_dat |> 
  mutate(year = as.factor(year),
         date_sampled = ymd(date_sampled))
```

# Introduction 

[enter introduction l8r]

# General Data Summaries  

## Study Period 

I.E., what dates/years are covered by the time series data? 

```{r}
range(study_dat$date_sampled)
```

We have data spanning Feb of 2004 to May of 2026. 



## Observations by Year 

Let's start by seeing how many measurements were recorded by year:

```{r}
study_dat |>  
  #mutate(year = as.numeric(year)) |> 
  group_by(year) |> 
  summarize(measurements_in_year = n()) |> 
  ggplot(aes(x = year, y = measurements_in_year)) +
  geom_point(size = 2) + 
  geom_line() +
  labs(
    title = "Number of Measurements by Year",
    x = "Year",
    y = "Count of Measurements"
  ) +
  theme_classic()

```

We have the most observations in 2024. We are also missing data for the year 2005.  


```{r}
summary(study_dat$year)
```
## Respirable silica by year 

Let's plot our respirable silica concentration measurements through our study period: 

```{r}

ggplot(data = study_dat, aes(x = year, y = silica_conc))+
  geom_boxplot()+
  labs(title = "Respirable Silica Measurements by Year", x = "Year", y = "Silica (mcg/m3)") +
  theme_classic()

#same, this time by date 

ggplot(data = study_dat, aes(x = date_sampled, y = silica_conc)) +
  geom_point(size=1, alpha = 0.5) +
  scale_x_date(date_breaks = "year", date_labels = "%b %Y") +
  labs(
    title = "Respirable Silica Measurements by Year",
    x = "Date",
    y = "Silica (mcg/m3)"
  )+ 
  theme_classic()+ 
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
``` 

Let's try adding in a smooth line to see if there are any apparent long-term patterns: 

```{r}
ggplot(data = study_dat, aes(x = date_sampled, y = silica_conc)) +
  geom_point(size=1, alpha = 0.5) + 
  geom_smooth()+
  scale_x_date(date_breaks = "year", date_labels = "%b %Y") +
  labs(
    title = "Respirable Silica Measurements by Year",
    x = "Date",
    y = "Silica (mcg/m3)"
  )+ 
  theme_classic()+ 
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

This may not be possible to see or use, as we may not have enough observations to appropriately fit our time series with year. We can try "zooming in": 

```{r} 
ggplot(data = study_dat, aes(x = date_sampled, y = silica_conc)) +
  geom_point(size=1, alpha = 0.5) + 
  geom_smooth()+
  scale_x_date(date_breaks = "year", date_labels = "%b %Y") +
  labs(
    title = "Respirable Silica Measurements by Year",
    x = "Date",
    y = "Silica (mcg/m3)"
  )+ 
  theme_classic()+ 
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
scale_y_continuous(
    limits = c(0, 500),
    breaks = seq(0, 500, by = 100)
  ) 
```

Interesting, it looks like there is no clear trend here. We could attempt a nonlinear model?  

Also, there does not appear to be any seasonality in our data. 


Let's add in our regulatory levels, and zoom in on our graph to see the patterns easier: 

```{r}
# Counting the observations outside the graphical limits
n_outside <- sum(study_dat$silica_conc > 1000, na.rm = TRUE) 

#Graph:

ggplot(data = study_dat, aes(x = date_sampled, y = silica_conc)) +
  geom_point(size = 1, alpha = 0.5) +
  
  #regulatory references
  geom_hline(aes(yintercept = 50, linetype = "PEL (50 mcg/m³)", color = "PEL"), 
             linewidth = 0.8, show.legend = TRUE) +
  geom_hline(aes(yintercept = 25, linetype = "Action Level (25 mcg/m³)", color = "Action Level"), 
             linewidth = 0.8, show.legend = TRUE) +
  
  # reducing scale of graphs
  scale_y_continuous(
    limits = c(0, 1000),
    breaks = seq(0, 1000, by = 200)
  ) +
  
  scale_x_date(date_breaks = "year", date_labels = "%b %Y") +
  
  # Coloring reference lines
  scale_color_manual(
    values = c("PEL" = "#d62728", "Action Level" = "#ff7f0e"),
    name = "Regulatory Limits"
  ) +
  scale_linetype_manual(
    values = c("PEL (50 mcg/m³)" = "solid", "Action Level (25 mcg/m³)" = "dashed"),
    name = "Regulatory Limits"
  ) +
  
  labs(
    title = "Respirable Silica Measurements by Year",
    x = "Date",
    y = "Silica (mcg/m³)",
    caption = paste0("Note: ", n_outside, " observation(s) exceed the y-axis limit of 1,000 mcg/m³")
  ) +
  
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
```

Let's investigate how our data differs depending on if it was taken from the IMIS code of 9000, or if we did the calculation to get the respirable silica concentration: 

```{r}

#version where we facet by IMIS code 
study_dat <- study_dat |> 
  mutate(imis = factor(imis, levels = c("9010_9130", "9000"),
                       labels = c("Combined Quartz and Respirable Dust", "Respirable Silica")))

ggplot(data = study_dat, aes(x = date_sampled, y = silica_conc)) +
  geom_point(size=1, alpha = 0.5) + 
  facet_grid(. ~ imis)+
  scale_x_date(date_breaks = "year", date_labels = "%b %Y") +
  labs(
    title = "Respirable Silica Measurements by Year",
    x = "Date",
    y = "Silica (mcg/m3)"
  )+ 
  theme_classic()+ 
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```  



Let's look at this a bit differently, with a bar graph that tells us how many observations exceed the PEL per year: 

```{r}
 study_dat <- study_dat |> 
  mutate(over_PEL = silica_conc >= 50) 

study_dat |>
  count(year, over_PEL) |>
  mutate(PEL_status = if_else(over_PEL, "Exceeds PEL", "Below PEL")) |>
  ggplot(aes(x = year, y = n, fill = PEL_status)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = n), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            size = 3) +
  scale_fill_manual(
    values = c("Exceeds PEL" = "#d62728", "Below PEL" = "#2ca02c"),
    name = "Status"
  ) +
  labs(
    title = "Respirable Silica Measurements: Exceeding vs. Below PEL (50 mcg/m³) by Year",
    x = "Year",
    y = "Count of Measurements"
  ) +
  theme_classic() +
  theme(legend.position = "top")
  
``` 

What about how many establishments are the same? 

```{r}
study_dat |> 
  group_by(establishment) |> 
  count() |> 
  filter(n>1)
``` 
Looks like there are 285 distinct establishments that have the same name. But something to consider is that our establishment sometimes has pretty similar naming conventions. These likely refer to the same institutions, just with some differences in terms of naming. We can look at this further: 

```{r} 
library(stringdist)
standardize_establishment <- function(x) {
  x |>
    str_squish() |>                          # trim + collapse internal whitespace
    str_to_upper() |>
    str_replace_all("&", "AND") |>           # unify ampersand vs "AND"
    str_replace_all("[[:punct:]]", "") |>    # strip periods, commas, hyphens, etc.
    str_squish()                             # re-squish in case punctuation removal left double spaces
} 
study_dat <- study_dat |> mutate(establishment = standardize_establishment(establishment)) 


study_dat |> 
  group_by(establishment) |> 
  count() |> 
  filter(n>1)
```

That helped reduce it down to 268, but we still see some examples where the spellings are inconsistent. We may need to be a bit more literal when we explore this by identifying the exact instances where establishments are mislabeled.  

```{r}
study_dat |> 
  group_by(establishment) |> 
  count()
```

Going through the list, FORTUNATELY only one establishment appears to be the same: 
1. "ALL GRANITE AND MARBLE CORP" and "ALL GRANITE AND MARBLE CORPORATION"  

Let's look at how the establishments change the distribution of silica:

```{r}
multiple_obs <- study_dat |> 
  group_by(establishment) |> 
  count() |> 
  filter(n>1) |> 
  pull(establishment)

study_dat |> 
  filter(establishment %in% multiple_obs) |> 
  group_by(establishment) |> 
  summarize(mean = round(mean(silica_conc), 2), sd = round(sd(silica_conc), 2))
```
We can see that there is a lot of individual variation in the data. We can also assume that some of these are being skewed by really dramatic outliers (take, for example, CL Granite and marble llc, whose mean silica is 1,376 with an sd of 2,478.07). 

# Sequential modeling 

Let's start fitting our time series model to this data. We will start with a simple model that looks at the association between year and the silica conc: 

```{r}
mod_linear_reg <- glm(silica_conc ~ year, data = study_dat) 
summary(mod_linear_reg)
```
 
That gave us the trend when year is a character variable. Since we are missing 2005, we may need to initalize our data at year 2005. Then we can create a year counter for the year in the study, and estimate a numeric result: 

### Subsetting data to exclude year 2005 

```{r}
dat_06 <- study_dat |> 
  filter(date_sampled > "2005-12-31") |> 
  mutate(year = as.numeric(year))
``` 

Doing so has reduced our sample size down to 1,043.  

Let's try our glm again: 
```{r}
mod_linear_reg <- glm(silica_conc ~ year, data = dat_06)
summary(mod_linear_reg)
```

Very interesting, this time we see a small reduction in concentration over the years, but this is VERY insignificant. 

Let's proceed investigating this, adding in our random effect for establishment to establish whether there is within-group variation: 

```{r} 


coeftest(mod_linear_reg, 
         vcov = vcovCL(mod_linear_reg, 
                       cluster = ~ establishment,
                       type = "HC1")) 


``` 

Not much has changed at all!   



### Over PEL

What if we instead investigate a logistic model that investigates whether the number of facilities exceeding the PEL has changed over the years? 

```{r}
mod_logistic  <- glm(over_PEL~year, data = dat_06, 
                     family = binomial)
summary(mod_logistic)
``` 

We should not need to control for establishment here, since we are only looking at whether the proportion of facilities over PEL changed by year, but we can add in a random effect for establishment either way: 

```{r}

coeftest(mod_logistic,
         vcov = vcovCL(mod_logistic,
                       cluster = ~ establishment,
                       type = "HC1")) 
``` 

Looks like it is slightly less significant, but still there! Let's interpret our outcome more clearly:  

```{r}
mod_log_cluster <- coeftest(mod_logistic,
         vcov = vcovCL(mod_logistic,
                       cluster = ~ establishment,
                       type = "HC1"))   
mod_log_cluster

tidy_result <- tibble(
  term = rownames(mod_log_cluster),
  estimate = exp(mod_log_cluster[,"Estimate"]),
  std.error = mod_log_cluster[, "Std. Error"],
  statistic = mod_log_cluster[, "z value"],
  p.value = mod_log_cluster[, "Pr(>|z|)"]
) %>%
  mutate(
    conf.low = exp(log(estimate) - 1.96 * std.error),
    conf.high = exp(log(estimate) + 1.96 * std.error)
  ) %>%
  select(term, estimate, conf.low, conf.high, std.error, statistic, p.value)

tidy_result |> 
  mutate(across(where(is.numeric), ~round(., 3)))

``` 

It looks like we see a slight decrease in the odds of the measurements being above PEL annually, controlling for the variation of each individual establishment.  

### Over Action Level 

What about the odds of annual measurements exceeding the action level? 

```{r}
dat_06 <- dat_06 |> 
  mutate(over_action = silica_conc >= 25)   

mod_logistic2  <- glm(over_action~year, data = dat_06, 
                     family = binomial)
summary(mod_logistic2)

mod_log_cluster2 <- coeftest(mod_logistic2,
         vcov = vcovCL(mod_logistic2,
                       cluster = ~ establishment,
                       type = "HC1"))   
mod_log_cluster2

tidy_result2 <- tibble(
  term = rownames(mod_log_cluster2),
  estimate = exp(mod_log_cluster2[,"Estimate"]),
  std.error = mod_log_cluster2[, "Std. Error"],
  statistic = mod_log_cluster2[, "z value"],
  p.value = mod_log_cluster2[, "Pr(>|z|)"]
) %>%
  mutate(
    conf.low = exp(log(estimate) - 1.96 * std.error),
    conf.high = exp(log(estimate) + 1.96 * std.error)
  ) %>%
  select(term, estimate, conf.low, conf.high, std.error, statistic, p.value)

tidy_result2 |> 
  mutate(across(where(is.numeric), ~round(., 3)))

```
There is no significant change in measurements exceeding the action level by year. 
