---
title: "02_Exploration"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true 
    code_folding: hide 
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(janitor)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

study_dat <- read.csv("../CleanData/study_dat.csv") 

study_dat <- study_dat |> 
  mutate(year = as.factor(year),
         date_sampled = ymd(date_sampled))
```

# Introduction 

[enter introduction l8r]

# General Data Summaries 

## Observations by Year 

Let's start by seeing how many measurements were recorded by year:

```{r}
study_dat |>  
  #mutate(year = as.numeric(year)) |> 
  group_by(year) |> 
  summarize(measurements_in_year = n()) |> 
  ggplot(aes(x = year, y = measurements_in_year)) +
  geom_point(size = 2) + 
  geom_line() +
  labs(
    title = "Number of Measurements by Year",
    x = "Year",
    y = "Count of Measurements"
  ) +
  theme_classic()

```

```{r}

ggplot(data = study_dat, aes(x = year, y = silica_conc))+
  geom_boxplot()+
  labs(title = "Respirable Silica Measurements by Year", x = "Year", y = "Silica (mcg/m3)") 

#same, this time by date 

ggplot(data = study_dat, aes(x = date_sampled, y = silica_conc)) +
  geom_point(size=1, alpha = 0.5) +
  scale_x_date(date_breaks = "year", date_labels = "%b %Y") +
  labs(
    title = "Respirable Silica Measurements by Year",
    x = "Date",
    y = "Silica (mcg/m3)"
  )+ 
  theme_classic()+ 
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
``` 
Let's try adding in a smooth line to see if there are any apparent long-term patterns: 

```{r}
ggplot(data = study_dat, aes(x = date_sampled, y = silica_conc)) +
  geom_point(size=1, alpha = 0.5) + 
  geom_smooth()+
  scale_x_date(date_breaks = "year", date_labels = "%b %Y") +
  labs(
    title = "Respirable Silica Measurements by Year",
    x = "Date",
    y = "Silica (mcg/m3)"
  )+ 
  theme_classic()+ 
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
This may not be possible to see or use, as we may not have enough observations to appropriately fit our time series with year. 



This is kind of hard to interpret. Let's add in our regulatory levels, and zoom in on our graph to see the patterns easier: 

```{r}
# Counting the observations outside the graphical limits
n_outside <- sum(study_dat$silica_conc > 1000, na.rm = TRUE) 

#Graph:

ggplot(data = study_dat, aes(x = date_sampled, y = silica_conc)) +
  geom_point(size = 1, alpha = 0.5) +
  
  #regulatory references
  geom_hline(aes(yintercept = 50, linetype = "PEL (50 mcg/m³)", color = "PEL"), 
             linewidth = 0.8, show.legend = TRUE) +
  geom_hline(aes(yintercept = 25, linetype = "Action Level (25 mcg/m³)", color = "Action Level"), 
             linewidth = 0.8, show.legend = TRUE) +
  
  # reducing scale of graphs
  scale_y_continuous(
    limits = c(0, 1000),
    breaks = seq(0, 1000, by = 200)
  ) +
  
  scale_x_date(date_breaks = "year", date_labels = "%b %Y") +
  
  # Coloring reference lines
  scale_color_manual(
    values = c("PEL" = "#d62728", "Action Level" = "#ff7f0e"),
    name = "Regulatory Limits"
  ) +
  scale_linetype_manual(
    values = c("PEL (50 mcg/m³)" = "solid", "Action Level (25 mcg/m³)" = "dashed"),
    name = "Regulatory Limits"
  ) +
  
  labs(
    title = "Respirable Silica Measurements by Year",
    x = "Date",
    y = "Silica (mcg/m³)",
    caption = paste0("Note: ", n_outside, " observation(s) exceed the y-axis limit of 1,000 mcg/m³")
  ) +
  
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
```

Let's investigate how our data differs depending on if it was taken from the IMIS code of 9000, or if we did the calculation to get the respirable silica concentration: 

```{r}

#version where we facet by IMIS code 
study_dat <- study_dat |> 
  mutate(imis = factor(imis, levels = c("9010_9130", "9000"),
                       labels = c("Combined Quartz and Respirable Dust", "Respirable Silica")))

ggplot(data = study_dat, aes(x = date_sampled, y = silica_conc)) +
  geom_point(size=1, alpha = 0.5) + 
  facet_grid(. ~ imis)+
  scale_x_date(date_breaks = "year", date_labels = "%b %Y") +
  labs(
    title = "Respirable Silica Measurements by Year",
    x = "Date",
    y = "Silica (mcg/m3)"
  )+ 
  theme_classic()+ 
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```  



Let's look at this a bit differently, with a bar graph that tells us how many observations exceed the PEL per year: 

```{r}
study_dat <- study_dat |> 
  mutate(over_PEL = silica_conc >= 50) 

study_dat |>
  count(year, over_PEL) |>
  mutate(PEL_status = if_else(over_PEL, "Exceeds PEL", "Below PEL")) |>
  ggplot(aes(x = year, y = n, fill = PEL_status)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = n), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            size = 3) +
  scale_fill_manual(
    values = c("Exceeds PEL" = "#d62728", "Below PEL" = "#2ca02c"),
    name = "Status"
  ) +
  labs(
    title = "Respirable Silica Measurements: Exceeding vs. Below PEL (50 mcg/m³) by Year",
    x = "Year",
    y = "Count of Measurements"
  ) +
  theme_classic() +
  theme(legend.position = "top")
  
```
