---
title: "AIHA Abstract Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true 
    code_folding: hide 
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(janitor) 
library(viridis) 
library(sandwich)
library(lmtest) 
library(broom) 
library(equatiomatic)


# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

study_dat <- read.csv("../CleanData/study_dat.csv") 

study_dat <- study_dat |> 
  mutate(year = as.factor(year),
         date_sampled = ymd(date_sampled))
```

# Introduction 

The following data exploration will produce descriptive statistics and explore the distribution of data in terms of duration of exposure for use of an AIHA abstract. 

# Data Exclusion 

For the purposes of this analysis, we will begin by excluding observations taking place in 2026 so we can have full year data: 

```{r}
study_dat |> 
  filter(year == 2026) |> 
  count() 

study_dat <- study_dat |> 
  filter(year != 2026)
```
We excluded 92 observations.


# General Data Summaries   

How many observations? 
```{r}
nrow(study_dat)
```

## Study Period 

I.E., what dates/years are covered by the time series data? 

```{r}
range(study_dat$date_sampled)
```
  
  
## How many samples per year? 

```{r}
study_dat |> 
  group_by(year) |> 
  summarize(n_year = n()) |> 
  summarize(
    min_n = min(n_year),
    max_n = max(n_year)
  ) 

study_dat |> 
  group_by(year) |> 
  summarize(n_year = n()) 
  
```
 
## Duration 

### N by duration

```{r}
study_dat |> 
  group_by(duration) |> 
  count()
```
### Concentration by Duration 

```{r}
study_dat |> 
  group_by(duration) |> 
  summarize(
    min_conc = min(silica_conc, na.rm = TRUE),
    max_conc = round(max(silica_conc, na.rm = TRUE),1),
    n_total  = n(), 
    below_lod = sum(lod), 
    pct_lod = round((100 * below_lod / n_total),1),
    over_AL  = sum(silica_conc > 25, na.rm = TRUE),
    over_PEL = sum(silica_conc > 50, na.rm = TRUE),
    pct_over_AL  = round((100 * over_AL / n_total),1),
    pct_over_PEL = round((100 * over_PEL / n_total),1)
  )
```
 
## LOD  

How many observations were LOD? By duration?  

```{r}
study_dat |> 
  group_by(lod) |> 
  count() 

#percentage 
(74/2330)*100 

#by duration 
study_dat |> 
  group_by(duration,lod) |> 
  count() 
``` 

## Over PEL  

Let's start by summarizing in terms of by year: 

```{r} 
study_dat <- study_dat |> 
  mutate(over_PEL = silica_conc >= 50) 

over_PEL_annual <- study_dat |> 
  group_by(year) |> 
  summarize(annual_pct_over_pel = round(mean(over_PEL == "TRUE")*100, 2))

over_PEL_annual 
```

Next, let's summarize in terms of duration category:  

```{r}
over_PEL_duration <- study_dat |> 
  group_by(duration) |> 
  summarize(annual_pct_over_pel = round(mean(over_PEL == "TRUE")*100, 2))

over_PEL_duration 
```


# Plots 

## AL and PEL plot 

```{r}
# --- Prep: order duration groups, label lod status ---
plot_dat <- study_dat %>%
  mutate(
    duration = factor(duration,
                         levels = c("Task-Based", "Short-Term", "Medium-Term", "Long-Term")),
    lod_status = factor(lod, levels = c(FALSE, TRUE),
                         labels = c("Reported Measurement", "Below LOD"))
  ) 

ggplot(plot_dat, aes(x = duration, y = silica_conc, shape = lod_status, color = lod_status)) +
  
  # Reported Measurement: drawn first (bottom layer), semi-transparent
  geom_jitter(
    data = subset(plot_dat, lod_status == "Reported Measurement"),
    width = 0.15, size = 1.3, alpha = 0.8
  ) +
  
  # Below LOD: drawn second (top layer), fully opaque, no alpha
  geom_jitter(
    data = subset(plot_dat, lod_status == "Below LOD"),
    width = 0.15, size = 1.3, alpha = 1
  ) +

  # Reference limits (fixed colors via override so they don't clash with the point color scale)
  geom_hline(aes(yintercept = 25, linetype = "Action Level (25 µg/m³)"),
             color = "#ff7f0e", linewidth = 0.8) +
  geom_hline(aes(yintercept = 50, linetype = "PEL (50 µg/m³)"),
             color = "#d62728", linewidth = 0.8) +
  scale_y_log10(
    breaks = c(1, 5, 10, 25, 50, 100, 500, 1000, 5000), 
    limits = c(1, 6000)
  ) +
  scale_shape_manual(
    name = "Measurement Type",
    values = c("Reported Measurement" = 16,
               "Below LOD" = 17)
  ) +
  scale_color_manual(
    name = "Measurement Type",
    values = c("Reported Measurement" = "grey40",
               "Below LOD" = "black")
  ) +
  scale_linetype_manual(
    name = "Reference Limits",
    values = c("Action Level (25 µg/m³)" = "dashed",
               "PEL (50 µg/m³)" = "dotdash")
  ) +
  guides(
    linetype = guide_legend(override.aes = list(colour = c("#ff7f0e", "#d62728")))
  ) +
  labs(
    title = "Respirable Crystalline Silica Concentration by Sample Duration Category",
    x = NULL,
    y = "RCS Concentration (µg/m³) [Log10 Scale]"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    legend.box = "vertical", 
    legend.key.size = unit(0.3, "cm"),
     legend.text = element_text(size = 8),
    legend.title = element_text(size = 9)
  )
``` 


## Over PEL Plot: 

```{r, fig.width = 10} 
study_dat <- study_dat |> 
  mutate(over_PEL = silica_conc > 50) 

over_PEL <- study_dat |> 
  group_by(year) |> 
  summarize(annual_pct_over_pel = round(mean(over_PEL == "TRUE")*100, 2))

over_PEL 
ggplot(over_PEL, aes(year, annual_pct_over_pel))+
  geom_col()+ 
  labs(x = "Year", y = "Percentage of Measurements Exceeding 50 mcg/m³ (%)")+
  ggtitle("Percentage of OSHA Respirable Silica Concentration Measurements\n Exceeding the PEL", subtitle = "By Year" )+ 
  
 geom_text(aes(label = paste0(annual_pct_over_pel, "%"), vjust = -0.5), size = 2.5)+  
  
   theme_classic()+
  
  theme(plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5), 
    axis.text.x = element_text(angle = 45, hjust = 1))
```

What about Over PEL by Duration?? 

```{r, fig.width = 12, fig.height= 8}
over_PEL_duration <- study_dat |> 
  group_by(year,duration) |> 
  summarize(annual_pct_over_pel = round(mean(over_PEL == "TRUE")*100, 2))

over_PEL_duration 
ggplot(over_PEL_duration, aes(year, annual_pct_over_pel))+ 
  facet_wrap(~duration)+
  geom_col()+ 
  labs(x = "Year", y = "Percentage of Measurements Exceeding 50 mcg/m³ (%)")+
  ggtitle("Percentage of OSHA Respirable Silica Concentration Measurements\n Exceeding the PEL", subtitle = "By Year and Duration" )+ 
  
 geom_text(aes(label = paste0(annual_pct_over_pel, "%"), vjust = -0.5), size = 2)+  
  
   theme_classic()+
  
  theme(plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5), 
    axis.text.x = element_text(angle = 45, hjust = 1))
```


# Temporal Trends 

## Median concentration reported per year 

The OSHA citations database was used to analyze the median concentration reported per year. We can replicate the same analysis:  

Do we have all of the years covered by this data? 

```{r}
study_dat |> 
  select(year) |> 
  unique()
``` 

We are missing data for the year 2005. We will also have to do the year counter process.

## Year Counter Creaqtion

```{r}
dat_clean <- study_dat |>  
  mutate(year = as.numeric(as.character(year))) |>  
  arrange(year) |>
  mutate(year_counter = dense_rank(year) - 1) 

dat_clean |> 
  distinct(year, year_counter) |> 
  arrange(year)
``` 

## Crude mod:  

Get the median by year first, then fit linear trend.

```{r}
median_by_year <- dat_clean |> 
  group_by(year_counter) |> 
  summarize(median_conc = median(silica_conc, na.rm = TRUE))  

ggplot(data=median_by_year,aes(year_counter,median_conc))+ 
  geom_point()+
  geom_line()

mod_median_trend <- lm(median_conc ~ year_counter, data = median_by_year)
summary(mod_median_trend)
```
There is a statistically significant relationship here whereby the median concentration increases by 1.3 units for each additional year in the study period.  


## Median by Duration 

Get the median by year and duration:

```{r}
median_by_year_and_duration <- dat_clean |> 
  group_by(year_counter,duration) |> 
  summarize(median_conc = median(silica_conc, na.rm = TRUE))  

ggplot(data=median_by_year_and_duration,aes(year_counter,median_conc))+  
  facet_wrap(~duration)+
  geom_point()+
  geom_line()

mod_median_trend2 <- lm(median_conc ~ year_counter*duration, data = median_by_year_and_duration)
summary(mod_median_trend2)
```


There is no significant trend when we differentiate by duration.


# All measurements 

## Linear Model

Let's try our glm again, this time with the pure silica conc instead of median concentration: 
```{r}
mod_linear_reg <- glm(silica_conc ~ year_counter, data = dat_clean)
summary(mod_linear_reg) 
extract_eq(mod_linear_reg)
```

Let's control for our within-establishment variation: 

```{r}
coeftest(mod_linear_reg, 
         vcov = vcovCL(mod_linear_reg, 
                       cluster = ~ establishment,
                       type = "HC1")) 


```

### By Duration 

This time, let's see whether duration interacts with this relationship: 

```{r}
mod_linear_reg2 <- glm(silica_conc ~ year_counter * duration, data = dat_clean)
summary(mod_linear_reg2) 
extract_eq(mod_linear_reg2)
```
This time it appears that as years increase the concentration goes down, but when it is task based the conc. intercept increases... 

Let's plot what the model is showing us: 

```{r}
# Generate predictions from the model across the range of year_counter for each duration
pred_grid <- dat_clean |> 
  distinct(duration) |> 
  tidyr::crossing(year_counter = seq(min(dat_clean$year_counter, na.rm = TRUE), 
                                       max(dat_clean$year_counter, na.rm = TRUE), 
                                       length.out = 100))

pred_grid$predicted <- predict(mod_linear_reg2 , newdata = pred_grid)

ggplot() +
  geom_point(data = dat_clean, aes(x = year_counter, y = silica_conc, color = duration), alpha = 0.3) +
  geom_line(data = pred_grid, aes(x = year_counter, y = predicted, color = duration), linewidth = 1.2) +
  facet_wrap(~ duration, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none")

```


### Over PEL

What if we instead investigate a logistic model that investigates whether the number of facilities exceeding the PEL has changed over the years? 

```{r} 



mod_logistic  <- glm(over_PEL~year_counter, data = dat_clean, 
                     family = binomial)
summary(mod_logistic) 
extract_eq(mod_logistic)
``` 

No significant relationship here. 

We should not need to control for establishment here, since we are only looking at whether the proportion of facilities over PEL changed by year, but we can add in a random effect for establishment either way: 

```{r}

coeftest(mod_logistic,
         vcov = vcovCL(mod_logistic,
                       cluster = ~ establishment,
                       type = "HC1")) 
```  



# Do Establishments improve over time? 

We want to check out the establishments that have multiple observations, and investigate whether they improve over time: 


Some establishments have multiple samples on the same day, so we will take the median conc only for those instances: 

```{r}
est_daily <- study_dat |> 
  group_by(establishment, date_sampled) |> 
  summarize(median_conc = median(silica_conc, na.rm = TRUE), .groups = "drop")
``` 

Next we only keep establishments with multiple distinct sampling dates: 

```{r}
est_multi <- est_daily |> 
  group_by(establishment) |> #just looking at establishment
  filter(n() >= 2) |> #only those that have multiple measurements
  arrange(establishment, date_sampled)#in order of date sampled
``` 


Let's figure out the change between consecutive measurements: 

```{r}
est_changes <- est_multi |> 
  group_by(establishment) |> 
  mutate(
    conc_change = median_conc - lag(median_conc), #median conc measurement MINUS the previous one
    days_elapsed = as.numeric(date_sampled - lag(date_sampled)), #number of days between consecuitve measurements
    years_elapsed = days_elapsed / 365.25, #converting days into years
    rate_per_year = conc_change / years_elapsed #the change in establishments divided by time elaspsed
  ) |> 
  filter(!is.na(conc_change))  #removing rows where conc_change is NA (each establishment's first row -- only retaining a measured change between two timepoints)
```

```{r}
est_changes

est_changes |> 
  ungroup() |> 
  summarize(
    mean_rate = mean(rate_per_year, na.rm = TRUE),
    median_rate = median(rate_per_year, na.rm = TRUE)
  )

#NOTE: this includes changes between the same establishment at different time points (i.e., if an establishment was measured 3 times, it will contribute 2 changes below)
est_changes |> 
  ungroup() |> 
  mutate(direction = case_when(
    conc_change < 0 ~ "improved",
    conc_change > 0 ~ "worsened",
    TRUE ~ "unchanged"
  )) |> 
  count(direction) |> 
  mutate(pct = 100 * n / sum(n))
```


```{r, fig.height=20, fig.width=20} 
ggplot(est_multi, aes(x = date_sampled, y = median_conc)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ establishment, scales = "free_x") +
  labs(x = "Date sampled", y = "Median silica concentration") +
  theme_minimal() +
  theme(strip.text = element_text(size = 7))

```

