---
title: "AIHA Abstract Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true 
    code_folding: hide 
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(janitor) 
library(viridis) 
library(sandwich)
library(lmtest) 
library(broom) 
library(equatiomatic)


# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

study_dat <- read.csv("../CleanData/study_dat.csv") 

study_dat <- study_dat |> 
  mutate(year = as.factor(year),
         date_sampled = ymd(date_sampled))
```

# Introduction 

The following data exploration will produce descriptive statistics and explore the distribution of data in terms of duration of exposure for use of an AIHA abstract. 

# Data Exclusion 

For the purposes of this analysis, we will begin by excluding observations taking place in 2026 so we can have full year data: 

```{r}
study_dat |> 
  filter(year == 2026) |> 
  count() 

study_dat <- study_dat |> 
  filter(year != 2026)
```
We excluded 92 observations.


# General Data Summaries   

How many observations? 
```{r}
nrow(study_dat)
```

## Study Period 

I.E., what dates/years are covered by the time series data? 

```{r}
range(study_dat$date_sampled)
```
 
 
## LOD  

How many observations were LOD? By duration?  

```{r}
study_dat |> 
  group_by(lod) |> 
  count() 

#percentage 
(74/2330)*100 

#by duration 
study_dat |> 
  group_by(duration,lod) |> 
  count() 
``` 

## Over PEL  

Let's start by summarizing in terms of by year: 

```{r} 
study_dat <- study_dat |> 
  mutate(over_PEL = silica_conc >= 50) 

over_PEL_annual <- study_dat |> 
  group_by(year) |> 
  summarize(annual_pct_over_pel = round(mean(over_PEL == "TRUE")*100, 2))

over_PEL_annual 
```

Next, let's summarize in terms of duration category:  

```{r}
over_PEL_duration <- study_dat |> 
  group_by(duration) |> 
  summarize(annual_pct_over_pel = round(mean(over_PEL == "TRUE")*100, 2))

over_PEL_duration 
```


# Plots 

## AL and PEL plot 

```{r}
# --- Prep: order duration groups, label lod status ---
plot_dat <- study_dat %>%
  mutate(
    duration = factor(duration,
                         levels = c("Task-Based", "Short-Term", "Medium-Term", "Long-Term")),
    lod_status = factor(lod, levels = c(FALSE, TRUE),
                         labels = c("Reported Measurement", "Below LOD"))
  ) 

ggplot(plot_dat, aes(x = duration, y = silica_conc, shape = lod_status, color = lod_status)) +
  
  # Reported Measurement: drawn first (bottom layer), semi-transparent
  geom_jitter(
    data = subset(plot_dat, lod_status == "Reported Measurement"),
    width = 0.15, size = 1.3, alpha = 0.8
  ) +
  
  # Below LOD: drawn second (top layer), fully opaque, no alpha
  geom_jitter(
    data = subset(plot_dat, lod_status == "Below LOD"),
    width = 0.15, size = 1.3, alpha = 1
  ) +

  # Reference limits (fixed colors via override so they don't clash with the point color scale)
  geom_hline(aes(yintercept = 25, linetype = "Action Level (25 µg/m³)"),
             color = "#ff7f0e", linewidth = 0.8) +
  geom_hline(aes(yintercept = 50, linetype = "PEL (50 µg/m³)"),
             color = "#d62728", linewidth = 0.8) +
  scale_y_log10(
    breaks = c(1, 5, 10, 25, 50, 100, 500, 1000, 5000), 
    limits = c(1, 6000)
  ) +
  scale_shape_manual(
    name = "Measurement Type",
    values = c("Reported Measurement" = 16,
               "Below LOD" = 17)
  ) +
  scale_color_manual(
    name = "Measurement Type",
    values = c("Reported Measurement" = "grey40",
               "Below LOD" = "black")
  ) +
  scale_linetype_manual(
    name = "Reference Limits",
    values = c("Action Level (25 µg/m³)" = "dashed",
               "PEL (50 µg/m³)" = "dotdash")
  ) +
  guides(
    linetype = guide_legend(override.aes = list(colour = c("#ff7f0e", "#d62728")))
  ) +
  labs(
    title = "Respirable Crystalline Silica Concentration by Sample Duration Category",
    x = NULL,
    y = "RCS Concentration (µg/m³) [Log10 Scale]"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    legend.box = "vertical", 
    legend.key.size = unit(0.3, "cm"),
     legend.text = element_text(size = 8),
    legend.title = element_text(size = 9)
  )
``` 



